Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies

### 2014-02-06 paper/Abstract.tex
A pass through end of Introduction

%Correct implementation of cache coherent systems is of high importance in
%industrial use.
We provide a precise formal definition of a cache coherent system as one
which obeys store atomicity. We give a Coq-based proof of a directory-based MSI cache
coherence protocol for an infinite family of cache hierarchies. We specify a
set of invariants which imply store atomicity, enabling other protocols to be
verified for store atomicity automatically as long as they satisfy the
invariants. Our key insight for proving these invariants is a set of
fundamental properties which are shared with other protocols like MESI, MOSI
and MOESI, enabling reuse of portions of our MSI proof for their verification.

%Previously, it has been shown
%that these atomic transitions can be translated automatically into efficient
%hardware. Verifying the specification based on atomic transitions therefore
%guarantees the correctness of the actual implementation.
%The proof is built in a modular fashion, identifying key invariants common with
%other protocols like MESI, MOSI, \etc. We informally show how these other
%protocols can be proved reusing modules from the proof of the MSI protocol.

>>>>

Cache-coherent shared memory is the foundation for today's high-performance
multicore hardware platforms.  Previous work has verified correctness of
shared memory implementations for finite-state systems, for fixed topologies of
processors and caches.  We present the first formal verification of correctness
for an \emph{infinite} family of memory hierarchies, via machine-checked proofs
in the Coq proof assistant.  First, we identify a set of general invariants
and prove formally that they imply cache coherence.  Then we give a proof
for the concrete example of the directory-based MSI protocol, establishing
correct behavior for arbitrary tree-shaped memory hierarchies.
Our key insight for proving the invariants is a set of
fundamental properties shared with other protocols like MESI, MOSI,
and MOESI, enabling reuse of portions of our MSI proof for their verification.

### 2014-02-06 paper/Introduction.tex
A pass through end of Introduction

Formal verification of cache coherence protocols has traditionally been done
using model-checking based approaches. This signficantly restricts the size as
well as complexity of the system that has to be verified; most model-checking
based approaches can not handle more than one level of hierarchy. To the best
of our knowledge, the largest system that has been verified using
model-checking based approaches has only 12 processors. \xxx{citation:
http://www.cs.utah.edu/formal\_verification/pdf/xiaofang\_dissertation.pdf (used
data from their related work must cite)}.

Apart from scalability concerns, another problem with model-checking based
approaches is that it provides no intuition as to why a protocol is correct, if
the model checker successfully proves a protocol. Developing a new protocol
therefore therefore can not use any intuition that could have been gained from
the first protocol. Moreover the new protocol has to be verified from scratch.

We propose a theorem-prover based approach for proving cache coherence
protocols. Using this approach eliminates the scalability and complexity
problems faced by model-checkers. But, the downside is that considerable
laborious effort is needed by verification people to provide a proof for the
protocol, and the complexity of the protocol directly translates into effort.
Our motivating question in this paper is,
\emph{can we modularize formal proofs of cache coherence protocols so that
verification experts do the hard work once and for all, and then hardware
designers can instantiate the results to different designs with much less
effort?}

While we have not reached a push-button state in our work, we do develop a
modular proof strategy, where portions of our proof can be used \emph{as is} to
prove other related but distinct protocols. Moreover, the proof gives solid
insights on how the protocol works correctly, which can be used to design other
protocols.

The main contributions of this paper are as follows:
\begin{enumerate}
\item \emph{We give a precise formal definition of a cache coherent system as one
which obeys store atomicity. The definition is abstract in that it does not
rely on any hardware implementation.}
\item \emph{We provide a Coq-based proof of the directory-based MSI cache coherence
protocol with inclusive caches, and no silent drops.}
\item \emph{We present a set of invariants which are not restricted to the MSI
protocol, but would imply store atomicity. Thus, other protocols will also
satisfy store atomicity as long as they obey the invariants.}
\item \emph{We present what we believe to be the key insight behind the
correctness of the MSI protocol. We believe that this property is true also of
other protocols like MESI, MOSI and MOESI, enabling reuse of portions of our
MSI proof for their verification.}
\end{enumerate}
%In practice, verification proceeds in
%two steps. First, a cache coherence protocol is constructed and verified. This
%protocol generally directly represents many of the key concurrent aspects of an
%implementation (\eg{} reads and writes to the system are broken into local
%atomic micro-operations) but ignores implementation details (\eg{} buffer
%sizing and memory access times). Second, once the protocol has been verified,
%it is handed to hardware experts who fill in the missing implementation details
%and realize it at the Register Transfer Level (RTL) of gates, wires, and
%registers, via further breaking down of local micro-operations to make them
%realizable in hardware. This effort is not a one-time translation, as
%performance properties are not easily estimated; a significant amount of
%performance testing is requires to select a good implementation. 
%
%\adamc{I don't understand what this paragraph is saying.}
%From a methodological perspective, this two-step process leaves much to be
%desired. The significant effort of verification is effectively lost by the
%concurrent non-trivial tasks of translation from the protocol representation to
%a hardware description and the necessary refinement to add missing
%implementation details to the protocol.

Protocols, post verification, are often described via Term Rewrite Systems
(TRS) \adamc{citation for TRSes?}, which apply equational logic to describe
systems at an interesting point on the spectrum between declarative and
operational methods.  It was shown by Dave
\etal{}~\cite{DNA:CoherenceImplementation} that such a protocol, after its
verification, can be transliterated into Bluespec~\cite{Bluespec:TFRG}, a
hardware description language based on Guarded Atomic Actions, a restriction of
TRS~\cite{Hoe:TCAD}. Bluespec specification can be directly implemented into
efficient hardware. In this way one reduces the correctness of the
implementation to the correctness of the protocol, with no need for separate
reasoning about an implementation. While the current Bluespec compiler does not
support some implementation elements that are common in practice (\eg{}
fairness guarantees, atomic actions running over multiple clock cycles), recent
work has provided key insights toward lifting the
restrictions~\cite{Karczmarek}. We believe that Bluespec and its future
evolutions represent a very promising platform for productive implementation of
hardware components, including cache coherent shared memories, starting from a
description meant for making verification easier. Our model of the MSI protocol
in Coq is described as an inductively defined set of transitions which is
directly translatable into Guarded Atomic Actions. In fact, our Coq-based model
is a direct transliteration of the MSI protocol specified in Bluespec.  This
ensures that the final hardware synthesized from the specification is formally
verified.

@@

\noindent\textbf{Paper Outline}. In Section~\ref{Background} we give a
formal definition of cache coherent system as one which obeys store atomicity.
We describe memory subsystems as mathematical functions and define store
atomicity as a property of the input/output pairs of the memory subsystem
function. We describe our MSI protocol and its representation as a
transition relation in Coq in Section~\ref{sec:System}. In
Section~\ref{safety} we discuss the structuring of our proof. We
present the invariants that are sufficient to prove store atomicity and give
the key insight that enables these invariants to be proved.
%We discuss some refinements in
%Section~\ref{Sec:Refinements} and conclude with a general discussion
%of the applicability of our technique.

>>>>

Formal verification of cache coherence protocols has traditionally been done
using model checking. That family of analyses significantly restricts the size as
well as complexity of the system that has to be verified; most model-checking
based approaches can not handle more than one level of hierarchy. To the best
of our knowledge, the largest system that has been verified using
model checking has only 12 processors, in an era when many observers are
predicting mainstream systems with hundreds of processor cores.
\xxx{citation:
http://www.cs.utah.edu/formal\_verification/pdf/xiaofang\_dissertation.pdf (used
data from their related work must cite)}.

Apart from scalability concerns, another downside of model checking
is that it provides no intuition as to why a protocol is correct, if
the model checker successfully proves a protocol. Developing a new protocol
therefore cannot use any intuition that could have been gained from
the first protocol. Moreover, the new protocol has to be verified from scratch.

We propose a theorem-prover based approach for proving cache coherence
protocols. Using this approach eliminates the scalability and complexity
problems faced by model checkers, but the downside is that considerable
laborious effort is needed by verification experts to provide a proof for the
protocol, and the complexity of the protocol directly translates into effort.
Our motivating question in this paper is,
\emph{can we modularize formal proofs of cache coherence protocols so that
verification experts do the hard work once and for all, and then hardware
designers can instantiate the results to different designs with much less
effort?}

While our present approach is not push-button, we have developed a
modular proof strategy, where portions of our proof can be used \emph{as is} to
prove other related but distinct protocols. Moreover, the proof gives
insights on why the protocol works correctly, which can be used to design other
protocols.

The main contribution of this paper is \textbf{the first verification of a
cache coherence protocol that is parametric over an infinite family of
memory hierarchies}.  Our proof is implemented in the Coq proof assistant and
checked via a very small trusted code base.  We also want to draw attention to
a few key elements of our methodology that should be of interest for future
verifications of this kind, both through literal code reuse and through insights
about key invariants that should generalize across protocols.
\begin{enumerate}
\item We give a \emph{precise formal definition of a cache-coherent system} as one
that obeys \emph{store atomicity}. The definition is abstract in that it does not
rely on any hardware implementation.
\item We provide a \emph{Coq-based proof of the directory-based MSI cache coherence
protocol} with inclusive caches and no silent drops.
\item We present \emph{a set of invariants} that are not restricted to the MSI
protocol but which imply store atomicity. Thus, other protocols will also
satisfy store atomicity as long as they obey the invariants.
\item We present what we believe to be \emph{the key insight behind the
correctness of the MSI protocol}. We believe that this property is true also of
other protocols like MESI, MOSI, and MOESI, enabling reuse of portions of our
MSI proof for their verification.
\end{enumerate}
%In practice, verification proceeds in
%two steps. First, a cache coherence protocol is constructed and verified. This
%protocol generally directly represents many of the key concurrent aspects of an
%implementation (\eg{} reads and writes to the system are broken into local
%atomic micro-operations) but ignores implementation details (\eg{} buffer
%sizing and memory access times). Second, once the protocol has been verified,
%it is handed to hardware experts who fill in the missing implementation details
%and realize it at the Register Transfer Level (RTL) of gates, wires, and
%registers, via further breaking down of local micro-operations to make them
%realizable in hardware. This effort is not a one-time translation, as
%performance properties are not easily estimated; a significant amount of
%performance testing is requires to select a good implementation. 
%
%\adamc{I don't understand what this paragraph is saying.}
%From a methodological perspective, this two-step process leaves much to be
%desired. The significant effort of verification is effectively lost by the
%concurrent non-trivial tasks of translation from the protocol representation to
%a hardware description and the necessary refinement to add missing
%implementation details to the protocol.

Protocols, post verification, are often described via Term Rewrite Systems
(TRS) \adamc{citation for TRSes?}, which apply equational logic to describe
systems at an interesting point on the spectrum between declarative and
operational methods.  It was shown by Dave
\etal{}~\cite{DNA:CoherenceImplementation} that such a protocol, after its
verification, can be transliterated into Bluespec~\cite{Bluespec:TFRG}, a
hardware description language based on Guarded Atomic Actions, a restriction of
TRS~\cite{Hoe:TCAD}. Bluespec specifications can be directly implemented into
efficient hardware. In this way one reduces the correctness of the
implementation to the correctness of the protocol, with no need for separate
reasoning about an implementation. While the current Bluespec compiler does not
support some implementation elements that are common in practice (\eg{}
fairness guarantees, atomic actions running over multiple clock cycles), recent
work has provided key insights toward lifting the
restrictions~\cite{Karczmarek}. We believe that Bluespec and its future
evolutions represent a very promising platform for productive implementation of
hardware components, including cache-coherent shared memories, starting from a
description designed to support verification. Our model of the MSI protocol
in Coq is described as an inductively defined set of transitions that is
directly translatable into Guarded Atomic Actions. In fact, our Coq-based model
is a direct transliteration of the MSI protocol specified in Bluespec,
ensuring that the final hardware synthesized from the specification is formally
verified.

@@

\noindent\textbf{Paper Outline}. In Section~\ref{Background} we give a
formal definition of a cache-coherent system as one that obeys store atomicity.
We describe memory subsystems as mathematical functions and define store
atomicity as a property of the input/output pairs of the memory subsystem
function. We describe our MSI protocol and its representation as a
transition relation in Coq in Section~\ref{sec:System}. In
Section~\ref{safety} we discuss the structuring of our proof. We
present the invariants that are sufficient to prove store atomicity and give
the key insight that enables these invariants to be proved.
%We discuss some refinements in
%Section~\ref{Sec:Refinements} and conclude with a general discussion
%of the applicability of our technique.

### 2014-02-06 paper/Title.tex
intuition
this the the correct version
A pass through end of Introduction

\title{Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof}

@@

\centerline{\footnotesize {\bf Keywords}: Cache coherence, theorem prover, cache hierarchy, guarded atomic actions}

>>>>

\title{Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies}

@@

\centerline{\footnotesize {\bf Keywords}: Cache coherence, proof assistant, cache hierarchy, guarded atomic actions}
